---
title: "PENDA: PErsoNalized Data Analysis"
subtitle: "Performing personalized data analysis with `penda`"
author: "Magali Richard, Florent Chuffart, Clémentine Decamps, Daniel Jost"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo=FALSE, eval=TRUE}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", fig.width=9, fig.height=6, eval=TRUE, echo=FALSE, results="hide", dpi=75)
layout(1, respect=TRUE)
penda::draw_penda()
```


# Introduction

`penda` (**PE**rso**N**alized **D**ifferential **A**nalysis ) is an open-access analysis software (R package). The purpose of this tutorial is to present a general method allowing to detect variation in gene expression for each tested sample compared to a reference panel.  

# Dataset

The dataset used for this example corresponds to the transcriptomes (normalized count, using DESeq2 method) of 44 patients with lung adenocarcinoma (PMID:25079552), including control and pathological samples. 

`data_ctrl` is a data matrix containing normalized counts for each control sample. 
The rownames of the matrix correspond to the gene_symbol, the colnames indicate the sample ID.

```{r echo=TRUE}
data_ctrl = penda::penda_data_ctrl
```

```{r results="verbatim"}
if (exists("nb_genes")) {
  set.seed(1)
  idx_genes = sample(rownames(data_ctrl), nb_genes)
} else {
  idx_genes = rownames(data_ctrl)
}
# idx_ctrl = sample(colnames(data_ctrl), 20)
idx_ctrl = colnames(data_ctrl)
data_ctrl = data_ctrl[idx_genes,idx_ctrl]

head(data_ctrl[,1:3])
dim(data_ctrl)
```

`data_case` is a data matrix containing normalized counts for each tumor sample. 
The rownames of the matrix correspond to the gene_symbol, the colnames indicate the sample ID.

```{r echo=TRUE}
data_case = penda::penda_data_case
```

```{r results="verbatim"}
#idx_case = sample(colnames(data_case), 30)
# idx_case = colnames(data_case)
data_case = data_case[rownames(data_ctrl),]
head(data_case[,1:3])
dim(data_case)
```


# Method

`penda` performs a 3-steps analysis: 

1. Data filtering and creation of the dataset

2. Gene ranking

3. Differential expression testing

This is a tutorial describing how to use `penda` package.

## Data filtering

The function `make_dataset` makes 3 steps to prepare the data at the analysis.  <br/>
First, it calls the function `detct_na_value` of Penda to remove samples with more than threshold % probes or patients with NA.  <br/>
Second, it calls the function `detect_zero_value` of Penda to remove genes with very low expression. You can choose the threshold of low expression, or by default it uses the function `normalmixEM` to compute this threshold on all the data normalized with a pseudo-log.  <br/>
Finally, the function ranks genes by the median in controls. This step is essential for the proper functioning of Penda.

```{r label="penda::make_dataset"}
if (!exists("Penda_dataset")) Penda_dataset = penda::make_dataset(data_ctrl, data_case, detectlowvalue = TRUE, detectNA = TRUE, threshold = 0.99)
data_ctrl_f = Penda_dataset$data_ctrl
data_case_f = Penda_dataset$data_case

dim(data_ctrl_f)
dim(data_case_f)
```

## Gene ranking

Second, we rank the gene expression using normal tissue sample (i.e. `the data_ctrl_f matrix`)

We used the function `compute_lower_and_higher_lists` to obtain a list of two matrices. 

The matrix **L** corresponds to the list of genes (i.e id) with a lower expression for each considered genes (lines). <br/>
The matrix **H** corresponds to the list of genes (i.e id) with a higher expression for each considered genes (lines).

For each, the list of **L** genes (resp. **H** genes) corresponds to the closest genes (up to `s_max` genes) that have a lower (resp. higher) expression in at least `threshold` % of the control samples. 

```{r, results="verbatim"}
threshold_LH = 0.99
s_max = 50
if (!exists("L_H_list")) L_H_list = penda::compute_lower_and_higher_lists(data_ctrl_f, threshold = threshold_LH, s_max = s_max)

layout(matrix(1:2, 1), respect=TRUE)
hist(rowSums(L_H_list$L != 0), xlab = "nb of L genes")
hist(rowSums(L_H_list$H != 0), xlab = "nb of H genes")
```

## Differential expression testing

Then, we used  `penda_test` to make the differential expresion test. This function analyses case samples one by one and compares their genes rank to the L_H_list.

Results are in the form of two matrix `$down_genes` and `$up_genes`, with for each case (column), TRUE if the gene (line) is deregulated (resp. down-regulated or up-regulated). 

```{r, results="verbatim"}
controls =  data_ctrl_f
samples = data_case_f
threshold = 0.9 
iterations =  20 
L_H_list =  L_H_list 
quant_test =  0
factor_test = 1.4

penda_res = penda::penda_test(samples = samples, 
                  controls = controls,
                   threshold = threshold, 
                   iterations =  iterations, 
                   L_H_list =  L_H_list, 
                   quant_test =  quant_test,
                   factor_test = factor_test)

colSums(penda_res$down_genes)
colSums(penda_res$up_genes)
```

```{r, echo = FALSE}
generate_data_bypatient = function(D_list, U_list){
  down = colSums(D_list)
  up = colSums(U_list)
  total = down + up
  patient_names = colnames(D_list)
  patients = rep(factor(patient_names , levels = patient_names [order(total)]), 3)
  variable = c(rep("down", length(patient_names)),
               rep("up", length(patient_names)),
               rep("total", length(patient_names)))
  value = c(down, up, total)
  return(data.frame(patients = patients,
                    variable = variable,
                    value = value))
}

data_bypatient = generate_data_bypatient(D_list = penda_res$down_genes,
                                         U_list = penda_res$up_genes)

plot_figure = function(data_patients){
library(ggplot2)
mytheme <- theme(panel.background = element_blank(),
  panel.grid.major = element_line(colour="black", size = (0.1)),
                 panel.grid.minor = element_blank())

p1 = ggplot(data_patients, aes(x = patients, y = value)) + 
    geom_line(aes(group = variable), colour = "grey80") + # theme_minimal(base_size = 20) +
    mytheme + ylim(0, 10000) + ylab("Number of  genes") + xlab("Patients") +
theme(axis.text.x = element_blank()) + geom_point(aes(colour = variable), 
    size = 0.5) + ggtitle("Degree of deregulation per patient") +  
scale_y_continuous(breaks = seq(0, 10000, 2000)) + scale_x_discrete(breaks = NULL) +
   scale_colour_manual(name  ="Gene deregulation", 
                      values = c("green", "darkblue", "red"), 
                       labels=c("DOWN", "UP & DOWN", "UP")) 
 return(p1)
}

plot_figure(data_patients = data_bypatient)

```

```{r fig.width=9, fig.height=9, eval = FALSE, echo = FALSE}

plot_heatmap_hclust = function (data) {
  sum(apply(is.na(data), 1, any))
  data = data[!apply(is.na(data), 1, any), ]

  # clustering base on correlation for tissues
  tmp_d = data
  tmp_d = t(tmp_d) - apply(tmp_d, 2, mean)
  tmp_d = t(tmp_d)
  tmp_d = cor(tmp_d, method="pe")
  dim(tmp_d)
  hc_col = hclust(dist(1 - tmp_d), method="complete")

  Colv = as.dendrogram(hc_col)
  dendrogram="col"      

  # clustering base on eucl. dist. for genes
  d = dist(data)
  hc_row = hclust(d, method="complete")
  Rowv = as.dendrogram(hc_row)
  dendrogram="both"      

  # col
  colors=c("blue", "gray", "red")
  cols = colorRampPalette(colors)(20)

  foo = gplots::heatmap.2(data, Rowv=Rowv, Colv=Colv, dendrogram="col", trace="none", col=cols,
                          labRow = FALSE,labCol = FALSE,
                          main=paste0("Penda (", nrow(data), " genes x ", ncol(data), " samples)"), mar=c(10,5), useRaster=TRUE)
}


plot_heatmap_hclust(data = penda_res$down_genes - penda_res$up_genes)
```


# Material and methods

The Penda vignette of the penda package version 1.0 was executed on `r floor(Penda_dataset$info["init_nb_genes"])`  genes, using `r floor(Penda_dataset$info["init_nb_ctrls"])` control samples and `r floor(Penda_dataset$info["init_nb_cases"])` case samples.

The data set was pretreated as following: 
`r floor(Penda_dataset$info["nb_genes_NA"])` genes and 
`r floor(Penda_dataset$info["nb_patients_NA"])` patients was removed for NA values, 
and `r floor(Penda_dataset$info["nb_genes_0"])` genes was removed for low value 
(under the threshold of `r Penda_dataset$info["val_min"]`). 
Genes and patients removed were under the conditions in `r Penda_dataset$info["threshold"]*100` % of cases.

`r ncol(controls)` controls were used to generate L and H lists using the following parameters: threshold = `r threshold_LH` and s_max = `r s_max`.

The penda method was then applied on `r ncol(data_case)` cases. With the following set of parameters: quantile = `r quant_test`, factor = `r factor_test` and threshold = `r threshold`.




# Session Information

```{r, results="verbatim"}
sessionInfo()
```



