---
title: "Presentation of Penda"
author: "Clémentine Decamps"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
# source("./R/simu.R")
# source("./R/ranking.R")
# source("./R/testing.R")

```



##Data processing

We import the control data, and we 150 gene for the 98 patients.
The two last patients are used to simulate deregulation data.
The function "detect_zero_value" allows to eliminate the gene with more than 80% of expression equal to 0.
The processed data has 104 genes.
```{r, tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=55)}
ctrl_data = readRDS('~/projects/perso_DA/data_ctrl.rds')[1:150,]
simu_data = ctrl_data[,97:98]
ctrl_data = ctrl_data[,-(97:98)]

null_values = detect_zero_value(ctrl_data, simu_data, 0.8)
ctrl_data = ctrl_data[!null_values,]
simu_data = simu_data[!null_values,]

```


##Ranking in control

The function "find_D_U_ctrl" returns two matrix. For each gene in columns, the row gene is 
"TRUE" if is **D**own-expresed or **U**p-expressed.

```{r, tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=55)}

D_U_ctrl = find_D_U_ctrl(ctrl_data, 0.01, 30, 0.8)
table(D_U_ctrl$D)
table(D_U_ctrl$U)

```
