---
title: "vignette_simulation"
author: "Clémentine Decamps"
date: "9 mars 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("~/projects/penda/R/simu.R")
source("~/projects/penda/R/ranking.R")
source("~/projects/penda/R/testing.R")

```

##Data processing

We import the control data, and we take 1000 genes for the 98 patients.
The five last patients are used to simulate deregulation data.
The function "detect_zero_value" was used before to eliminates genes with more than 99% of expression value under 10.
```{r, tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=55)}
ctrl_data = readRDS('~/projects/perso_DA/data_ctrl_process.rds')

simu_data = ctrl_data[,94:98]
colnames(simu_data) = c("P1","P2","P3","P4","P5")

ctrl_data = ctrl_data[,-(94:98)]

#Paramètres simuv1 : 30% des données, seuil à 60
simuv1 = readRDS("simuv1.rds")
simuv2 = readRDS("simuv2.rds")


```

##Ranking in control

The function "find_D_U_ctrl" returns two matrix. For each gene in columns, the row gene is 
"TRUE" if is **D**own-expresed or **U**p-expressed.

```{r, tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=55)}

# D_U_ctrl = find_D_U_ctrl(ctrl_data, 0.001, 4, 0.99)
# table(D_U_ctrl$D)
# table(D_U_ctrl$U)
# saveRDS(D_U_ctrl, "D_U_93p.rds")
D_U_ctrl = readRDS("D_U_93p.rds")

check_D_U (D_U_ctrl, ctrl_data, 0.99)

```


```{r, tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=55)}

#DU_rearragement permet de sortir deux jolies matrices D et U 
DU_rearrangement = function(multiple_tests, simu_data, threshold_values){
  #Définition des matrices D et U pour toutes les conditions
  D_simu = matrix(data = NA, nrow = length(simu_data)
                  , ncol = length(threshold_values)
                  , dimnames = list(names(simu_data), threshold_values)) 
  U_simu = matrix(data = NA, nrow = length(simu_data)
                  , ncol = length(threshold_values)
                  , dimnames = list(names(simu_data), threshold_values))
  
  #On réarrange D et U dans les matrices pour toutes les conditions
  for (r in 1:length(threshold_values)){
    D_simu[,r] = unlist(multiple_tests[(r*2-1)])
    U_simu[,r] = unlist(multiple_tests[r*2])
  }
  return(list(D = D_simu, U = U_simu))
}


#Mutiple D_U permet de faire le test pour plusieurs conditions de threshold
multiple_DU = function(ctrl_data, D_U_ctrl, simulation, threshold_values, iterations, quantile){
  #Test pour toutes les valeurs de threshold
  multiple_test = sapply(threshold_values, function(t){
    patient_test(ctrl_data, simulation$simulated_data, quantile, iterations, D_U_ctrl, t)
  })
  sorted_test = DU_rearrangement(multiple_test, simulation$initial_data, threshold_values )
  return (sorted_test)
} 


#Fais le test, puis le FDR
total = function(ctrl_data, D_U_ctrl, threshold_values, iterations, quantile, simulation){
  results = c()
  #Pour chaque patient
  for(p in 1:ncol(simulation$initial_data)){
    simulation_p = list(initial_data = simulation$initial_data[,p], simulated_data = simulation$simulated_data[,p]) 
      #On fait le test pour toutes les conditions
    test = multiple_DU(ctrl_data, D_U_ctrl, simulation_p, threshold_values, iterations, quantile)
    #Pour chaque seuil du test
    for(thre in 1:ncol(test$D)){
      #On calcule FP, TP, FN, TN
      results_simu = results_simulation(test$D[,thre], test$U[,thre], simulation_p)
      #On calcule FDR et TPR
      FDR = results_simu$FP / (results_simu$TP + results_simu$FP)
      TPR  = results_simu$TP / (results_simu$TP + results_simu$FN)
      #On bind les résultats
      results = rbind(results, c(p, colnames(test$D)[thre], FDR, TPR))
    }
  }
  colnames(results) = c("patient", "threshold", "FDR", "TPR")
  return(results)
}

threshold_values = c(0.003, 0.05, 0.10, 0.15, 0.20, 0.25, 0.30, 0.35, 0.40, 0.45, 0.50)

t_debut_v2 = Sys.time()
FDRetc_v2 = total(ctrl_data, D_U_ctrl, threshold_values, 20, 0.03, simuv2)
saveRDS(FDRetc_v2, "FDRetc_v2.rds")

t_debut_v1 = Sys.time()
FDRetc_v1 = total(ctrl_data, D_U_ctrl, threshold_values, 20, 0.03, simuv1)
saveRDS(FDRetc_v1, "FDRetc_v1.rds")

t_debut_v1_sansl0 = Sys.time()
FDRetc_v1_sansl0 = total(ctrl_data, D_U_ctrl, threshold_values, 20, 0, simuv1)

t_debut_v2_sansl0 = Sys.time()
FDRetc_v2_sansl0 = total(ctrl_data, D_U_ctrl, threshold_values, 20, 0, simuv2)
t_fin = Sys.time()



dfv2 = data.frame(patient = FDRetc_v2[,1], threshold = FDRetc_v2[,2], FDR = as.numeric(FDRetc_v2[,3]), TPR = as.numeric(FDRetc_v2[,4])) 
dfv1 = data.frame(patient = FDRetc_v1[,1], threshold = FDRetc_v1[,2], FDR = as.numeric(FDRetc_v1[,3]), TPR = as.numeric(FDRetc_v1[,4])) 

g2 = ggplot(dfv2, aes(x = TPR, y = FDR, color = threshold, group = patient)) + geom_point() + geom_line()

g1 = ggplot(dfv1, aes(x = TPR, y = FDR, color = threshold, group = patient)) + geom_point() + geom_line()



DF2 = data.frame(l0 = factor(rep(c("Avec l0", "Sans l0"), each = 55)), 
                 patient = c(FDRetc_v2[,1], FDRetc_v2_sansl0[,1]),
                 threshold = c(FDRetc_v2[,2], FDRetc_v2_sansl0[,2]), 
                 FDR = c(as.numeric(FDRetc_v2[,3]), as.numeric(FDRetc_v2_sansl0[,3])), 
                 TPR = c(as.numeric(FDRetc_v2[,4]), as.numeric(FDRetc_v2_sansl0[,4]))) 



```

