---
title: "Performing PErsoNalized Data Analysis with `penda`"
author: "Magali Richard, Florent Chuffart, Clémentine Decamps, Daniel Jost"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---

```{r, echo=FALSE}
draw_penda = function(){
  library(ggplot2)
  tetey = c(4, 5, 5.8, 5.8, 5.4, 5, 4, 3, 1.5, 1.3, 1.5, 2.5, 4)
  tetex = c(1, 1.75, 2.9, 4, 4.8, 5.2, 5.8, 6, 5, 3.8, 2.3, 1.2, 1)
  orgy = c(4.5, 5.2, 6, 6.2, 5.8, 5, 4.5)
  orgx = c(1.35, 1, 1.2, 2.3, 2.9, 1.75, 1.35)
  ordy = c(5.8, 6.2, 6, 5.6, 4.8, 4.2, 5, 5.4, 5.8)
  ordx = c(4, 4.8, 5.3, 5.9, 6, 5.7, 5.2, 4.8, 4)

  ydx = c(3.9, 4.2, 4.8, 5, 4.8, 4.2, 3.8, 3.9)
  ydy = c(4, 4.4, 4.2, 4, 3.3, 3.2, 3.5, 4)

  ygx = c(2.5, 2.2, 2.5, 3.2, 3.5, 3.2, 2.5)
  ygy = c(3.2, 4, 4.3, 4.3, 4, 3.3, 3.2 )

  points = data.frame(x = c(2.8, 3.5, 4.3)
                    ,y = c(3.8, 2.8, 3.8))

  graph = ggplot() + geom_path(aes(tetex, tetey)) + geom_path(aes(orgx, orgy, size = 3)) + geom_path(aes(ordx, ordy, size = 3)) + geom_path(aes(ydx, ydy, size = 3)) + geom_path(aes(ygx, ygy, size = 3)) + geom_point(data = points, aes(x = x, y = y, size = 3)) +
    theme_bw() +
    theme(legend.position="none") +
    theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
     ggtitle("penda",subtitle = "PErsoNalized Data Analysis" ) +
  theme(plot.title = element_text(lineheight=10, face="bold", hjust = 0.5), plot.subtitle = element_text(lineheight=.8, face="italic", hjust = 0.5)) 
  
  return(graph)
}
draw_penda()
draw_penda_better = function(){
  tetey = c(4, 5, 5.8, 5.8, 5.4, 5, 4, 3, 1.5, 1.3, 1.5, 2.5, 4)
  tetex = c(1, 1.75, 2.9, 4, 4.8, 5.2, 5.8, 6, 5, 3.8, 2.3, 1.2, 1)
  orgy = c(4.5, 5.2, 6, 6.2, 5.8, 5, 4.5)
  orgx = c(1.35, 1, 1.2, 2.3, 2.9, 1.75, 1.35)
  ordy = c(5.8, 6.2, 6, 5.6, 4.8, 4.2, 5, 5.4, 5.8)
  ordx = c(4, 4.8, 5.3, 5.9, 6, 5.7, 5.2, 4.8, 4)

  ydx = c(3.9, 4.2, 4.8, 5, 4.8, 4.2, 3.8, 3.9)
  ydy = c(4, 4.4, 4.2, 4, 3.3, 3.2, 3.5, 4)

  ygx = c(2.5, 2.2, 2.5, 3.2, 3.5, 3.2, 2.5)
  ygy = c(3.2, 4, 4.3, 4.3, 4, 3.3, 3.2 )

  points = data.frame(x = c(2.8, 3.5, 4.3)
                    ,y = c(3.8, 2.8, 3.8))

  plot(0,0,col=0, xlim=c(0,7), ylim=c(0,7), xaxt="n", yaxt="n", xlab="", ylab="", main="penda\n4x4")
  grid()
  polygon(tetex, tetey)
  polygon(ordx, ordy, lwd=8)
  polygon(orgx, orgy, lwd=8)
  polygon(ydx, ydy, lwd=8)
  polygon(ygx, ygy, lwd=8)
  points(points, cex=3, pch=16)
}
draw_penda_better()
```


# Introduction

`penda` (**PE**rso**N**alized **D**ifferential **A**nalysis ) is an open-access analysis software (R package). The purpose of this tutorial is to present a general method allowing to detect variation in gene expression for each tested sample compared to a reference panel.  

# Dataset

The dataset used for this example correspond to the transcriptomes (normalized count, using DESeq2 method) of 44 patients with lung adenocarcinoma (PMID:25079552), including control and pathological samples. 

`data_ctrl` is a data matrix containing normalized counts for each control sample. 
The rownames of the matrix correspond the gene_symbol , the colnames indicate the sample ID.

```{r, results="verbatim"}
data_ctrl = penda::data_ctrl
head(data_ctrl[,1:5])
is.matrix(data_ctrl)
dim(data_ctrl)
```

`data_case` is a data matrix containing normalized counts for each tumor sample. 
The rownames of the matrix correspond the gene_symbol , the colnames indicate the sample ID.

```{r, results="verbatim"}
data_case = penda::data_case
head(data_case[,1:5])
is.matrix(data_case)
dim(data_case)
```


# Method

`penda` performs a 3-steps analysis: 

1. Data filtering

2. Gene ranking

3. Differential expression testing

This is a tutorial describing how to use `penda` package.

## Data filtering

First we filter the data to remove genes with very low expression.

```{r}
null_values = penda::detect_zero_value(data_ctrl, data_case, threshold = 0.99, min = 10)
data_ctrl_f = data_ctrl[!null_values,] #filtered matrix
data_case_f = data_case[!null_values,] #filtered matrix
dim(data_ctrl_f)
dim(data_case_f)
```

## Gene ranking

Second, we rank the gene expression using normal tissue sample (i.e. `the data_ctrl_f matrix`)

We used the function `compute_down_and_up_list` which a list of two matrices. 

The matrix **D** corresponds to the list (i.e true) of downregulated genes for each considered genes (columns).
The matrix **U** corresponds to the list (i.e true) of upregulated genes for each considered genes (columns).

For each, the list of **D** genes (resp. **U** genes) corresponds to the closest genes (up to `s_max` genes) that are down (resp. up) regulated in at least `threshold` % of the control samples. 



```{r, results="verbatim"}
if (!file.exists("tmp/D_U_list.rds")){
  controls = data_ctrl_f
  D_U_list = penda::compute_down_and_up_list(controls, threshold = 0.99, s_max = 50)
  saveRDS(D_U_list , "tmp/D_U_list.rds")
}
D_U_list = readRDS("tmp/D_U_list.rds")
hist(colSums(D_U_list$D))
```

## Differential expression testing



```{r, results="verbatim"}
controls =  data_ctrl_f
samples = data_case_f[,1:2] 
threshold = 0.03 
iterations =  20 
D_U_list=  D_U_list 
quant_test =  0.2
factor_test = 1

if (!file.exists("tmp/penda_res.rds")){
  penda_res = penda::penda_test(samples = samples, 
                        controls = controls,
                   threshold = threshold, 
                   iterations =  iterations, 
                   D_U_list =  D_U_list, 
                   quant_test =  quant_test,
                   factor_test = factor_test)
  saveRDS(penda_res , "tmp/penda_res.rds")
}
penda_res = readRDS("tmp/penda_res.rds")
colSums(penda_res$down_genes)
colSums(penda_res$up_genes)
```


## Plot results

Summarize the results

```{r}
generate_data_bypatient = function(D_list, U_list){
  down = colSums(D_list)
  up = colSums(U_list)
  total = down + up
  patient_names = colnames(D_list)
  patients = rep(factor(patient_names , levels = patient_names [order(total)]), 3)
  variable = c(rep("down", length(patient_names)),
               rep("up", length(patient_names)),
               rep("total", length(patient_names)))
  value = c(down, up, total)
  return(data.frame(patients = patients,
                    variable = variable,
                    value = value))
}

generate_data_bygene = function(D_list, U_list){
  
  #get_density <- function(x, y, n = 100) {
 # dens <- MASS::kde2d(x = x, y = y, n = n)
 # ix <- findInterval(x, dens$x)
 # iy <- findInterval(y, dens$y)
 # ii <- cbind(ix, iy)
 # return(dens$z[ii])
 # }
  
  down = rowSums(D_list)
  up = rowSums(U_list)
  total = down + up
  gene_name = rownames(D_list)
  pc_up = (up / dim(D_list)[2]) * 100
  pc_tot = (total /  dim(D_list)[2]) * 100
  #density = get_density(x = pc_up, y = pc_tot)
  return(data.frame(gene_name = gene_name,
                    up = up,
                    down = down,
                    total = total,
                    pc_up = pc_up,
                   # density = density,
                    pc_tot = pc_tot))
}

data_bypatient = generate_data_bypatient(D_list = penda_res$down_genes,
                                         U_list = penda_res$up_genes)


data_bygene = generate_data_bygene(D_list = penda_res$down_genes,
                                   U_list = penda_res$up_genes)
```


```{r}
plot_figure = function(data_patients, data_genes){
  
mytheme <- theme(panel.background = element_blank(),
  panel.grid.major = element_line(colour="black", size = (0.1)),
                 panel.grid.minor = element_blank())

p1 = ggplot(data_patients, aes(x = patients, y = value)) + 
    geom_line(aes(group = variable), colour = "grey80") + # theme_minimal(base_size = 20) +
    mytheme + ylim(0, 10000) + ylab("Number of  genes") + xlab("Patients") +
theme(axis.text.x = element_blank()) + geom_point(aes(colour = variable), 
    size = 0.5) + ggtitle("Degree of deregulation per patient") +  
scale_y_continuous(breaks = seq(0, 10000, 2000)) + scale_x_discrete(breaks = NULL) +
   scale_colour_manual(name  ="Gene deregulation", 
                      values = c("green", "darkblue", "red"), 
                       labels=c("DOWN", "UP & DOWN", "UP")) 
 

p2 = ggplot(data_genes, aes(x = pc_up,   y = pc_tot)) + geom_point(size = 0.2, alpha = 0.1) +# scale_color_viridis() + 
  ylab("% of patients DEregulated") + 
  xlab("% of patients UPregulated") +
 # geom_density2d(colour="black") + 
  ggtitle("Degree of deregulation per genes") + 
    theme_minimal()

return(gridExtra::grid.arrange(p1,p2))
}

plot_figure(data_patients = data_bypatient, 
            data_genes = data_bygene)

```























