---
title: "Performing PErsoNalized Data Analysis with `penda`"
author: "Magali Richard, Florent Chuffart, Clémentine Decamps, Daniel Jost"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---

```{r, echo=FALSE}
draw_penda = function(){
  library(ggplot2)
  tetey = c(4, 5, 5.8, 5.8, 5.4, 5, 4, 3, 1.5, 1.3, 1.5, 2.5, 4)
  tetex = c(1, 1.75, 2.9, 4, 4.8, 5.2, 5.8, 6, 5, 3.8, 2.3, 1.2, 1)
  orgy = c(4.5, 5.2, 6, 6.2, 5.8, 5, 4.5)
  orgx = c(1.35, 1, 1.2, 2.3, 2.9, 1.75, 1.35)
  ordy = c(5.8, 6.2, 6, 5.6, 4.8, 4.2, 5, 5.4, 5.8)
  ordx = c(4, 4.8, 5.3, 5.9, 6, 5.7, 5.2, 4.8, 4)

  ydx = c(3.9, 4.2, 4.8, 5, 4.8, 4.2, 3.8, 3.9)
  ydy = c(4, 4.4, 4.2, 4, 3.3, 3.2, 3.5, 4)

  ygx = c(2.5, 2.2, 2.5, 3.2, 3.5, 3.2, 2.5)
  ygy = c(3.2, 4, 4.3, 4.3, 4, 3.3, 3.2 )

  points = data.frame(x = c(2.8, 3.5, 4.3)
                    ,y = c(3.8, 2.8, 3.8))

  graph = ggplot() + geom_path(aes(tetex, tetey)) + geom_path(aes(orgx, orgy, size = 3)) + geom_path(aes(ordx, ordy, size = 3)) + geom_path(aes(ydx, ydy, size = 3)) + geom_path(aes(ygx, ygy, size = 3)) + geom_point(data = points, aes(x = x, y = y, size = 3)) +
    theme_bw() +
    theme(legend.position="none") +
    theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
     ggtitle("penda",subtitle = "PErsoNalized Data Analysis" ) +
  theme(plot.title = element_text(lineheight=10, face="bold", hjust = 0.5), plot.subtitle = element_text(lineheight=.8, face="italic", hjust = 0.5)) 
  
  return(graph)
}
draw_penda()
```


# Introduction

`penda` (**PE**rso**N**alized **D**ifferential **A**nalysis ) is an open-access analysis software (R package). The purpose of this tutorial is to present a general method allowing to detect variation in gene expression for each tested sample compared to a reference panel.  

# Dataset

The dataset used for this example correspond to the transcriptomes (normalized count, using DESeq2 method) of 44 patients with lung adenocarcinoma (PMID:25079552), including control and pathological samples. 

`data_ctrl` is a data matrix containing normalized counts for each control sample. 
The rownames of the matrix correspond the gene_symbol , the colnames indicate the sample ID.

```{r, results="verbatim"}
data_ctrl = penda::data_ctrl
head(data_ctrl[,1:5])
is.matrix(data_ctrl)
dim(data_ctrl)
```

`data_case` is a data matrix containing normalized counts for each tumor sample. 
The rownames of the matrix correspond the gene_symbol , the colnames indicate the sample ID.

```{r, results="verbatim"}
data_case = penda::data_case
head(data_case[,1:5])
is.matrix(data_case)
dim(data_case)
```


# Method

## Data filtering

## Gene ranking

## Differential expression testing




































##Data processing

We import the control data, and we take 1000 genes for the 98 patients.
The five last patients are used to  make the dataset of simulations.
The function "detect_zero_value" was used before to eliminates genes with more than 99% of expression value under 10.
```{r, tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=55), eval=FALSE}
ctrl_data = readRDS('~/projects/perso_DA/data_ctrl_process.rds')[1:1000,]
simu_data = ctrl_data[,94:98]
colnames(simu_data) = c("P1","P2","P3","P4","P5")

ctrl_data = ctrl_data[,-(94:98)]
```


##Ranking in control

We used the function "find_D_U_ctrl" which returns two matrix. For each gene in columns, the row gene is "TRUE" if is **D**own-expresed or **U**p-expressed.
The limit is between the quantile 0.001/4 and the quantile 0.999*4. 99% of patients must be in the conditions.
The function "check_D_U" checks if not exists genes both up and down expressed, and for each gene :
- if the maximal down gene is less expressed in 99% of controls.
- if the minimal up gene is more expressed in 99 % of controls.


```{r, tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=55), eval=FALSE}

D_U_ctrl = find_D_U_ctrl(ctrl_data, 0.001, 4, 0.99)
table(D_U_ctrl$D)
table(D_U_ctrl$U)

check_D_U (D_U_ctrl, ctrl_data, 0.99)

```

##Simulation of dysregulation of gene expression

We want to dysregul genes of 3 control patients in "simu_data" to simulate a cancer dysregulation.
The function "simplified_simulation" randomly dysreguls a given proportion of genes expressions. Here, we choose 30% of genes.
We do the simulation on 3 patients, and then we concatenate the results.

```{r, tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=55), eval=FALSE}

simup1 = simplified_simulation(simu_data[,1], 0.3)
simup2 = simplified_simulation(simu_data[,2], 0.3)
simup3 = simplified_simulation(simu_data[,3], 0.3)
simulation = list()

simulation$simulated_data = matrix(c(simup1$simulated_data, simup2$simulated_data, simup3$simulated_data), ncol=3, dimnames = list(rownames(simu_data)))
simulation$initial_data = matrix(c(simup1$initial_data, simup2$initial_data, simup3$initial_data), ncol=3, dimnames = list(rownames(simu_data)))
simulation$changes_idx = matrix(c(simup1$changes_idx, simup2$changes_idx, simup3$changes_idx), ncol=3)

table(simulation$initial_data != simulation$simulated_data)
length(simulation$changes_idx)
```


#Test of the dysregulation

We make the test for the 3 dysregulated patients. 
The quantile of l0 is 0.03, it means that genes at the extremities of expression distribution (under 3% or above 97%) are not used in the control list.
0.2 is the threshold to test if genes has changed.
20 is the maximum number of iterations allowed to stabilize the results.

Please be careful to have gene names as row names in your datas. 

```{r, tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=55), eval=FALSE}

D_matrix = simulation$simulated_data
U_matrix = simulation$simulated_data
t12 = Sys.time()
for(p in 1:3){
  print(c("Patient number",p))
  D_U = patient_test(ctrl_data, simulation$simulated_data[,p], 0.03, 20, D_U_ctrl, 0.2)
  D_matrix[,p] = unlist(D_U[1])
  U_matrix[,p] = unlist(D_U[2])
}
t22 = Sys.time()

#Number of predicted down-regulated genes
colSums(D_matrix)
#Number of predicted up-regulated genes
colSums(U_matrix)
```

#Representation

The function "results_simulation" computes the number of false positive, true positive, false negative and false positive results from the test and the original datas.
The function "draw_results" make an histogram with these results.

```{r, tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=55), eval=FALSE}

results = results_simulation(D_matrix, U_matrix, simulation)

draw_results(results)
```



##Complex simulation 

Complex_simulation used the real distribution of dyresgulation in our cancers to makes the simulation.
We do the simulation on 3 patients, and then we concatenate the results.

```{r, tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=55), eval=FALSE}

csimup1 = complex_simulation(simu_data[,1])
csimup2 = complex_simulation(simu_data[,2])
csimup3 = complex_simulation(simu_data[,3])
csimulation = list()

csimulation$simulated_data = matrix(c(csimup1$simulated_data, csimup2$simulated_data, csimup3$simulated_data), ncol=3, dimnames = list(rownames(simu_data)))
csimulation$initial_data = matrix(c(csimup1$initial_data, csimup2$initial_data, csimup3$initial_data), ncol=3, dimnames = list(rownames(simu_data)))
csimulation$changes_idx = list(csimup1$changes_idx, csimup2$changes_idx, csimup3$changes_idx)

table(csimulation$initial_data != csimulation$simulated_data)

```

#Test of the dysregulation

We make the test for the 3 patients with a complex simulation. 

Please be careful to have gene names as row names in your datas. 
```{r, tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=55), eval=FALSE}

D_matrixc = csimulation$simulated_data
U_matrixc = csimulation$simulated_data
for(p in 1:3){
  print(c("Patient number",p))
  D_U = patient_test(ctrl_data, csimulation$simulated_data[,p], 0.03, 20, D_U_ctrl, 0.2)
  D_matrixc[,p] = unlist(D_U[1])
  U_matrixc[,p] = unlist(D_U[2])
}

#Number of predicted down-regulated genes
colSums(D_matrixc)
#Number of predicted up-regulated genes
colSums(U_matrixc)
```

#Representation

The function "results_simulation" computes the number of false positive, true positive, false negative and false positive results from the test and the original datas.
The function "draw_results" make an histogram with these results.

```{r, tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=55), eval=FALSE}

resultsc = results_simulation(D_matrixc, U_matrixc, csimulation)

draw_results(resultsc)
```
cl <- makeCluster(8)
t1 = Sys.time()
p=patient_test(ctrl_data, simuv1$simulated_data[,1], 0.03, 20, D_U_ctrl, 0.1)
t2 = Sys.time()


